---
title: "Web facilities"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
fontsize: 11pt
bibliography: '`r system.file("REFERENCES.bib", package="gaawr2")`'
csl: '`r system.file("csl", "nature-genetics.csl", package = "gaawr2")`'
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Web facilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.height = 8,
  fig.path = "web/",
  fig.width = 8,
  collapse = TRUE,
  comment = "#>",
  dev = "CairoPNG")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pkgs <- c("httr","httpuv","jsonlite","plumber","Rsamtools")
has_pkg <- function(p) requireNamespace(p, quietly = TRUE)
missing <- pkgs[!vapply(pkgs, has_pkg, logical(1))]
if (length(missing)) {
  warning("Missing packages: ", paste(missing, collapse = ", "))
}
installed <- pkgs[vapply(pkgs, has_pkg, logical(1))]
invisible(suppressMessages(lapply(installed, require, character.only = TRUE)))
```

# Hello, world!

## Python

Perhaps this is the simplest way to start from the Linux command line, nevertheless with Python,

```python
python3 -m http.server 8000
firefox http://127.0.0.1:8000
```

which works with `python3 -m http.server` by default when the port number 8000 is available.

## R

Our focus here is R with the following script.

```{r httpuv-server, eval=FALSE}
httpuv::startServer("0.0.0.0", 8000, list(
  call = function(req) {
    list(
      status = 200L,
      headers = list("Content-Type" = "text/plain"),
      body = "Hello, world!"
    )
  })
)
```

Upon accessing `http://127.0.0.1:8000`, we see the message "Hello, world!".

# URL availability

This is an attempt to implement the following CRAN policy,

> Packages which use Internet resources should fail gracefully with an informative message if the resource is not available or has changed (and not give a check warning nor error).

```{r url}
check_url_availability <- function(url) {
# Send a GET request to the URL
  response <- tryCatch({
    httr::GET(url)
  }, error = function(e) {
# If there was an error, return FALSE
    return(NULL)
  })
# If response is NULL or status code is not 200, return FALSE
  if (is.null(response) || httr::status_code(response) != 200) {
    message(paste("Warning: The URL", url, "is not accessible. Please check the link."))
    return(FALSE)
  }
# If status code is 200, the URL is available
  message(paste("The URL", url, "is accessible."))
  return(TRUE)
}

url_to_check <- "http://www.zotero.org/styles/nature-genetics"
is_available <- check_url_availability(url_to_check)

if (is_available) {
  message("Using CSL as usual.")
} else {
  message("Using fallback to local CSL file instead.")
}
```

Below, we assume that our working directory is the source package,

# Data source

We envisage a server from which data can be obtained, and two R packages **plumber** and **httpuv** are considered here.

## A meta-analysis

The compressed data `IL.18R1-1.tbl.gz` is based on METAL @willer10, involving study-[1|2|3] as follows,

```
SEPARATOR TAB
COLUMNCOUNTING STRICT
CHROMOSOMELABEL CHR
POSITIONLABEL POS
CUSTOMVARIABLE N
LABEL N as N
TRACKPOSITIONS ON
AVERAGEFREQ ON
MINMAXFREQ ON
ADDFILTER AF1 >= 0.001
ADDFILTER AF1 <= 0.999
MARKERLABEL SNP
ALLELELABELS A1 A2
EFFECTLABEL BETA
PVALUELABEL P
WEIGHTLABEL N
FREQLABEL AF1
STDERRLABEL SE
SCHEME STDERR
EFFECT_PRINT_PRECISION 8
STDERR_PRINT_PRECISION 8
GENOMICCONTROL OFF
LOGPVALUE ON
OUTFILE IL18R1B_dr- .tbl
PROCESS study-1-IL18R1B.fastGWA.gz
PROCESS study-2-IL18R1B.fastGWA.gz
PROCESS study-3-IL18R1B.fastGWA.gz
PROCESS study-1-IL18R1B-chrX.fastGWA.gz
PROCESS study-2-IL18R1B-chrX.fastGWA.gz
PROCESS study-3-IL18R1B-chrX.fastGWA.gz
ANALYZE HETEROGENEITY
CLEAR
```

Assuming the script is named `IL.18R1.metal`, `IL.18R1-1.tbl.gz` is generated from **htslib**, <https://www.htslib.org/download/>, as follows,

```bash
export protein=IL.18R1
metal ${protein}.metal 2>&1 | \
tee ${protein}-1.tbl.log
cat <(head -1 ${protein}-1.tbl) \
    <(sed '1d' ${protein}-1.tbl | \
sort -k1,1n -k2,2n) | \
bgzip -f > ${protein}-1.tbl.gz
tabix -S1 -s1 -b2 -e2 -f ${protein}-1.tbl.gz
rm ${protein}-1.tbl
```

The bgzipped data allows for access by genomic region as shown below.

## plumber

Software library `libsodium`, <https://doc.libsodium.org/>, is required for its installation.

It is `an API generator in R`, which has been tested as follows.

```{r plumber, eval=FALSE}
get_data <- function(filename, region)
{
# query_result <- seqminer::tabix.read(filename, region)
  tbx <- Rsamtools::TabixFile(filename)
  query_result <- Rsamtools::scanTabix(tbx, param = region)[[1]]
  if (length(query_result) == 0) {
    return(data.frame())
  }
  hdr <- c("Chromosome", "Position",
           "MarkerName", "Allele1", "Allele2", "Freq1", "FreqSE", "MinFreq", "MaxFreq",
           "Effect", "StdErr", "logP",
           "Direction", "HetISq", "HetChiSq", "HetDf", "logHetP", "N")
  df <- read.table(text = paste(query_result, collapse = "\n"), sep = "\t", col.names = hdr,
                   stringsAsFactors = FALSE)
  return(df)
}

plbr <- plumber::Plumber$new()
plbr$handle("GET", "/tests", function(req, res) {
  protein <- req$args$protein
  region <- req$args$region
  if (is.null(protein) || is.null(region)) {
    res$status <- 400
    return(list(error = "Both 'protein' and 'region' must be provided"))
  }
  filename <- file.path("tests",paste0(protein,"-1.tbl.gz"))
  print(filename)
  if (!file.exists(filename)) {
    res$status <- 404
    return(list(error = paste("File for", protein, "not found")))
  }
  data <- get_data(filename, region)
  json_data <- jsonlite::toJSON(data, dataframe = "rows", na = "null")
  res$setHeader("Content-Type", "application/json")
  return(json_data)
})
options(width = 200)
filename <- file.path("tests","IL.18R1-1.tbl.gz")
region <- "2:102700000-103800000"
data <- get_data(filename, region)
head(data,1)
plbr$run(port = 8001)
```

Indeed we can see that the first line of data,

```
  Chromosome  Position         MarkerName Allele1 Allele2 Freq1 FreqSE MinFreq MaxFreq  Effect StdErr  logP   Direction HetISq HetChiSq HetDf logHetP     N
1          2 102700138 chr2:102700138_A_G       a       g 0.087 0.0207  0.0641  0.1376 -0.0566 0.0239 -1.75 -?-+n-?--n+   78.2   36.757     8  -4.894 12799
```

and

```
Running plumber API at http://127.0.0.1:8001
Running swagger Docs at http://127.0.0.1:8001/__docs__/
```

## Data access

### Browser/console

So we get query results in JSON format from

- **browser**: `http://localhost:8001/tests?protein=IL.18R1&region=2:102700000-103800000`
- **command line interface**: `curl "http://localhost:8001/tests?protein=IL.18R1&region=2:102700000-103800000"`

Additional work required to get output from `curl` to a tab-delimited data,

```bash
curl "http://localhost:8001/tests?protein=IL.18R1&region=2:102700000-103800000" | \
jq -r '.[0] |
   fromjson |
   .[] |
   [
     .Chromosome, .Position, .MarkerName, .Allele1, .Allele2, .Freq1,
     .Effect, .StdErr, .logP, .Direction, .HetISq, .HetChiSq, .HetDf, .logHetP, .N
   ] |
   @tsv'
```

where

1. **.[0]**: Access the first element in the outer array (the string containing the JSON).
2. **fromjson**: Parse the string into a JSON object.
3. **.[]**: Iterate over the array inside the parsed JSON.
4. **[ ... ]**: Create an array of the values needed, each corresponds to a column in the TSV output.
5. **@tsv**: Convert the array into tab-separated values.

Note also that only selected columns (as in 4) are kept. The simplest way to have the header is add it manually,

```bash
(
  echo "Chromosome|Position|MarkerName|Allele1|Allele2|Freq1|Effect|StdErr|logP|Direction|HetISq|HetChiSq|HetDf|logHetP|N" | \
  sed 's/|/\t/g'
  curl command as above.
)
```

### R

The query above is easily furnished with **curl**:

```{r curl, eval=FALSE}
tmp <- tempfile()
curl::curl_download("http://localhost:8001/tests?protein=IL.18R1&region=2:102700000-103800000", tmp)
df <- jsonlite::fromJSON(readLines(tmp)) |>
      jsonlite::fromJSON(flatten=TRUE) |>
      as.data.frame()
dim(df)
```

giving

```
[1] 4779   18
```

## httpuv

The package gives a somewhat more involved version as follows,

```{r httpuv, eval=FALSE}
dir.create("content/assets", recursive = TRUE)
dir.create("content/lib", recursive = TRUE)
s <- httpuv::startServer(
  host = "0.0.0.0", 
  port = 5000,
  app = list(
    call = function(req) {
      list(
        status = 200L,
        headers = list(
          'Content-Type' = 'text/html',
          'Access-Control-Allow-Origin' = '*',
          'Access-Control-Allow-Methods' = 'GET, POST, OPTIONS',
          'Access-Control-Allow-Headers' = 'Content-Type'
        ),
        body = "Hello world!"
      )
    },
    staticPaths = list(
      "/assets" = "content/assets/", # Static assets
      "/lib" = httpuv::staticPath("content/lib", indexhtml = FALSE),
      "/lib/dynamic" = httpuv::excludeStaticPath()
    ),
    staticPathOptions = httpuv::staticPathOptions(indexhtml = TRUE)
  )
)
cat("Server running at http://0.0.0.0:5000\n")
s$stop()
```

so mappings are created from `content/[assets, lib]` to `assets` and `lib`, while `httpuv::excludeStaticPath()` indicates that requests
to `/lib/dynamic` will not be served as static files but could be handled dynamically by the app logic.

# References
